---
title: "Beyond PCA: t-SNE and UMAP"
subtitle: "What if variation follows a curved path?"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: cosmo
    highlight: tango
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 8,
  fig.align = "center"
)

# Install these packages before class if needed:
# install.packages(c("remotes", "tidyverse", "patchwork", "plotly",
#                    "Rtsne", "umap", "viridis"))
# remotes::install_github("jlmelville/snedata")

library(tidyverse)
library(patchwork)
library(plotly)
library(snedata)
library(Rtsne)
library(umap)
library(viridis)
```

## Introduction

PCA finds **straight directions** that capture variation in a dataset. This works well when the structure of the data is approximately linear.

But what if the important variation follows a **curved path**?



t-SNE and UMAP take a different approach. Instead of finding new straight axes, they try to create a low-dimensional map in which observations that are similar in the original data tend to remain near one another.

Unlike PCA, t-SNE and UMAP do not produce axes that can be interpreted using variable loadings or variance explained. Instead, we mainly examine which observations are placed near one another. *IMPORTANT:The exact positions, shapes, and distances—especially between separate groups—should not be interpreted literally.*


### Learning goals

By the end of this exercise, you should be able to:

- explain why PCA can struggle with curved structure;
- describe the neighborhood idea behind t-SNE and UMAP;
- compare what PCA, t-SNE, and UMAP preserve or distort; and
- identify one caution when interpreting a t-SNE or UMAP plot.

---

## Part 1: The Swiss roll problem

A Swiss roll is a two-dimensional sheet curled through three-dimensional space. An ideal two-dimensional representation would **unroll the sheet** while keeping nearby positions along the sheet near one another.

```{r swiss_roll_creation}
set.seed(123)
n_points <- 1000

t <- runif(n_points, 0, 3 * pi)
x <- t * cos(t) + rnorm(n_points, 0, 0.1)
y <- t * sin(t) + rnorm(n_points, 0, 0.1)
z <- runif(n_points, 0, 5) + rnorm(n_points, 0, 0.1)

swiss_roll <- data.frame(x, y, z, t)

cat(
  "Points:", nrow(swiss_roll), "\n",
  "Measured dimensions:", ncol(swiss_roll) - 1, "\n"
)
```

### View the original data

```{r swiss_roll_3d}
plot_3d <- plot_ly(
  swiss_roll,
  x = ~x,
  y = ~y,
  z = ~z,
  color = ~t,
  colors = viridis_pal()(100),
  type = "scatter3d",
  mode = "markers",
  marker = list(size = 3, opacity = 0.7)
) %>%
  layout(
    title = "The Swiss Roll",
    scene = list(
      xaxis = list(title = "X"),
      yaxis = list(title = "Y"),
      zaxis = list(title = "Z")
    )
  )

plot_3d
```



---

## Part 2: Compare PCA, t-SNE, and UMAP

PCA looks for straight directions that capture variation. t-SNE and UMAP instead focus strongly on **neighborhoods**: groups of observations that are similar to one another in the original feature space.

t-SNE and UMAP follow the same broad idea:

- identify observations that are near one another in the original space;
- create a two-dimensional arrangement; and
- adjust the arrangement so that many of the original neighbors remain close.

The t-SNE and UMAP coordinates are generated by the algorithms. Their axes do not correspond to measured variables or directions of variance.

UMAP typically runs more efficiently/faster for large, biological-scale datasets such as single-cell sequencing. 

### Run all three methods on the Swiss roll

```{r swiss_roll_comparison}
# PCA
pca_swiss <- prcomp(swiss_roll[, 1:3], scale. = TRUE)

pca_swiss_data <- data.frame(
  Dim1 = pca_swiss$x[, 1],
  Dim2 = pca_swiss$x[, 2],
  t = swiss_roll$t
)

# t-SNE
set.seed(123)
tsne_swiss <- Rtsne(
  swiss_roll[, 1:3],
  dims = 2,
  perplexity = 30,
  verbose = FALSE
)

tsne_swiss_data <- data.frame(
  Dim1 = tsne_swiss$Y[, 1],
  Dim2 = tsne_swiss$Y[, 2],
  t = swiss_roll$t
)

# UMAP
set.seed(123)
umap_swiss <- umap(
  swiss_roll[, 1:3],
  n_neighbors = 15,
  min_dist = 0.1,
  verbose = FALSE
)

umap_swiss_data <- data.frame(
  Dim1 = umap_swiss$layout[, 1],
  Dim2 = umap_swiss$layout[, 2],
  t = swiss_roll$t
)

# Make the three plots
p_pca_swiss <- ggplot(pca_swiss_data, aes(Dim1, Dim2, color = t)) +
  geom_point(size = 2, alpha = 0.7) +
  scale_color_viridis_c(name = "Position\nalong roll") +
  labs(title = "PCA", x = "PC1", y = "PC2") +
  theme_minimal()

p_tsne_swiss <- ggplot(tsne_swiss_data, aes(Dim1, Dim2, color = t)) +
  geom_point(size = 2, alpha = 0.7) +
  scale_color_viridis_c(name = "Position\nalong roll") +
  labs(title = "t-SNE", x = "t-SNE 1", y = "t-SNE 2") +
  theme_minimal()

p_umap_swiss <- ggplot(umap_swiss_data, aes(Dim1, Dim2, color = t)) +
  geom_point(size = 2, alpha = 0.7) +
  scale_color_viridis_c(name = "Position\nalong roll") +
  labs(title = "UMAP", x = "UMAP 1", y = "UMAP 2") +
  theme_minimal()

# Display all three together
p_pca_swiss + p_tsne_swiss + p_umap_swiss +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

### Question 1

**Which method(s) unroll the Swiss roll most clearly?**

*Answer:tSNE 1 and Umap*


### Question 2

**Which method do you think best reveals the continuous path through the Swiss roll? Explain your reasoning.**

*Answer: t-sne more continuous *


### Parameter choices matter

t-SNE uses `perplexity` to help define the scale of a local neighborhood. 

UMAP has two commonly adjusted parameters:

- `n_neighbors`: the approximate neighborhood size;
- `min_dist`: how tightly points may pack together in the final map.

Changing these values can noticeably change the visualization. The map is an algorithmic representation, not a literal coordinate system.

```{r tsne_perplexity_comparison}
perplexities <- c(15, 60)

perplexity_data <- map_dfr(perplexities, function(p) {
  set.seed(123)
  fit <- Rtsne(
    swiss_roll[, 1:3],
    dims = 2,
    perplexity = p,
    verbose = FALSE
  )

  data.frame(
    tSNE1 = fit$Y[, 1],
    tSNE2 = fit$Y[, 2],
    t = swiss_roll$t,
    perplexity = paste("Perplexity =", p)
  )
})

ggplot(perplexity_data, aes(tSNE1, tSNE2, color = t)) +
  geom_point(size = 1.5, alpha = 0.7) +
  facet_wrap(~perplexity) +
  scale_color_viridis_c(name = "Position\nalong roll") +
  labs(title = "The t-SNE map changes with perplexity") +
  theme_minimal()
```

### Question 3

**How does changing the t-SNE perplexity from 15 to 60 change the resulting plot?**

*Answer: 15 disjoint, 60 more continuous spread of points*


### Question 4

**Why do you think increasing perplexity has this effect on the t-SNE visualization?**

*Answer: Local neighborhood increased in size?*

---

## Part 3: The mammoth exercise

The Swiss roll was constructed to contain a curved sheet. Now we will compare the three methods using a 3D scan of a mammoth skeleton.

A mammoth has only three measured dimensions, so we can see the original structure before reducing it. Biological datasets may have hundreds or thousands of dimensions, so we normally cannot make this direct comparison.

### Load and view the mammoth data

```{r mammoth_data}
mammoth3d <- download_mammoth10k()

cat(
  "Points:", nrow(mammoth3d), "\n",
  "Dimensions:", ncol(mammoth3d), "\n"
)

head(mammoth3d)
```

```{r mammoth_3d_plot}
mammoth_plotly <- plot_ly(
  mammoth3d,
  x = ~X,
  y = ~Y,
  z = ~Z,
  color = ~Z,
  colors = viridis_pal()(100),
  type = "scatter3d",
  mode = "markers",
  marker = list(size = 2, opacity = 0.6)
) %>%
  layout(
    title = "Original 3D Mammoth Scan",
    scene = list(
      xaxis = list(title = "X"),
      yaxis = list(title = "Y"),
      zaxis = list(title = "Z")
    )
  )

mammoth_plotly
```

The color records each point's original Z coordinate. We will carry it into every two-dimensional plot so that we can track where points came from.

### Question 5

**Looking at the original 3D mammoth, what features do you predict will be retained well by a straight two-dimensional projection such as PCA?**

*Answer: Side profile of  but not features like ribs*


### Question 6

**Which parts of the mammoth contain curved or branching structures that might be difficult to represent in two dimensions?**

*Answer:Ribs, legs, tusks*


### Question 7

**Do you expect a nonlinear method such as t-SNE or UMAP to produce a more recognizable mammoth than PCA? Why or why not?**

*Answer: Yes, more flexibility*

### Run PCA, t-SNE, and UMAP

NOTE: this may take a minute to run, and will slow knitting of the html

```{r mammoth_pca_tsne_umap}
# The three coordinate axes use the same units, so we center but do not rescale.
mammoth_pca <- prcomp(
  mammoth3d,
  center = TRUE,
  scale. = FALSE
)

set.seed(123)
mammoth_tsne <- Rtsne(
  mammoth3d,
  dims = 2,
  perplexity = 30,
  verbose = FALSE
)

set.seed(123)
mammoth_umap <- umap(
  mammoth3d,
  n_neighbors = 15,
  min_dist = 0.1,
  verbose = FALSE
)

pca_mammoth_data <- data.frame(
  Dim1 = mammoth_pca$x[, 1],
  Dim2 = mammoth_pca$x[, 2],
  original_z = mammoth3d$Z
)

tsne_mammoth_data <- data.frame(
  Dim1 = mammoth_tsne$Y[, 1],
  Dim2 = mammoth_tsne$Y[, 2],
  original_z = mammoth3d$Z
)

umap_mammoth_data <- data.frame(
  Dim1 = mammoth_umap$layout[, 1],
  Dim2 = mammoth_umap$layout[, 2],
  original_z = mammoth3d$Z
)

p_pca_mammoth <- ggplot(
  pca_mammoth_data,
  aes(Dim1, Dim2, color = original_z)
) +
  geom_point(size = 1, alpha = 0.6) +
  scale_color_viridis_c(name = "Original Z") +
  labs(title = "PCA", x = "PC1", y = "PC2") +
  theme_minimal()

p_tsne_mammoth <- ggplot(
  tsne_mammoth_data,
  aes(Dim1, Dim2, color = original_z)
) +
  geom_point(size = 1, alpha = 0.6) +
  scale_color_viridis_c(name = "Original Z") +
  labs(title = "t-SNE", x = "t-SNE 1", y = "t-SNE 2") +
  theme_minimal()

p_umap_mammoth <- ggplot(
  umap_mammoth_data,
  aes(Dim1, Dim2, color = original_z)
) +
  geom_point(size = 1, alpha = 0.6) +
  scale_color_viridis_c(name = "Original Z") +
  labs(title = "UMAP", x = "UMAP 1", y = "UMAP 2") +
  theme_minimal()

p_pca_mammoth + p_tsne_mammoth + p_umap_mammoth +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")
```

### Question 8

**Which method best preserves the mammoth's overall shape?**

*Answer: Umap yo*


### Question 9

**What local relationships or structures become easier to see with t-SNE or UMAP compared with PCA?**

*Answer: Bones, patterns of variation along curves*


### Question 10

**Which method, t-SNE or UMAP, better preserves the overall organization of the original mammoth—for example, keeping bones approximately connected to the body parts where they belong?**

*Answer: Umap, single cell sequencing, DIMENSIONS NOT INTERPRETABLE YO!*


---

## Final reflection

### Question 11

**In 3–5 sentences, explain why PCA may fail when important variation follows a curved path, how t-SNE or UMAP approaches this problem differently, and one reason to be cautious when interpreting distances, shapes, or clusters in a t-SNE or UMAP plot.**

*Answer:Linear vs non-linear *

