---
title: "Exercise - K-means Clustering - Human Population Genetics"
author: "Luis Alegría"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
editor_options:
  chunk_output_type: console
---

Load the packages and data. Install any missing packages

Download `phased_HGDP+India+Africa_2810SNPs-regions1to36.csv` from D2L and place it in the same folder as this R Markdown file.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(umap)

df <- read_csv("phased_HGDP+India+Africa_2810SNPs-regions1to36.csv")
```


## Learning goals

By the end of this exercise, you should be able to:

- explain the utility of clustering;
- run k-means clustering on a high-dimensional dataset;
- interpret a cluster assignment and a cluster centroid;
- examine how the choice of $K$ affects the results; and
- compare data-driven clusters with known biological labels.

## The dataset

Today, we will work with data from the **Human Genome Diversity Project (HGDP)**. The dataset contains genotypes for 1,107 individuals from 63 populations around the world, measured at approximately 2,800 single-nucleotide polymorphisms (SNPs).



SNPs are DNA positions in the genome where individuals vary. 
For example,
rs6696404 is the unique database identifier for one particular SNP:

-  rs means Reference SNP
- 6696404 is its identification number in NCBI dbSNP
- It is an A/G variant on chromosome 1
- Its genomic coordinate depends on the reference-genome version; on GRCh38 it is approximately chromosome 1 position 3,075,369.


- Each **row** represents one individual.
- Columns beginning with `rs` contain that individual's genotype at a SNP.
- Columns such as `pop` and `region` contain known population information. We will not use these labels to create the clusters; we will use them afterward to help interpret the results.


### Question 1

**What is each observation (row) in this dataset? What are the features that we will use for clustering? Why don't we want to include `region` and `pop` as clustering features?**

*Answer: Each observation, or row, in the HGDP represents an individual and contains SNP and population features. The features used for clustering are any column beginning with rs, as these represent a person's SNP genotype encoded by 0, 1, or 2. The region and pop features are not used because they are categorical, and the clustering algorithm requires numerical features to calculate distances.*


## Prepare the genotype data

We first extract the SNP columns and standardize each SNP to have a mean of 0 and a standard deviation of 1. This puts the SNPs on comparable scales.

After centering and scaling, a value of 0 represents the mean for a SNP. For this exercise, we therefore replace missing values with 0.

```{r prepare-data}
genotypes_scaled <- df %>%
  select(starts_with("rs")) %>%
  scale()

genotypes_scaled[is.na(genotypes_scaled)] <- 0
```

The original dataset has approximately 2,800 genotype dimensions. We will first use PCA to make a smaller representation containing 20 principal components. This reduces noise and makes the clustering faster while retaining major patterns of genetic variation.

This is not an essential step, however it will greatly speed up the clustering. 

```{r run-pca}
pca <- prcomp(genotypes_scaled, rank. = 20)
pca_scores <- pca$x
```

**PCA and UMAP represent the observations in a new coordinate system. Clustering does something different: it assigns the observations to groups.**

## Make a shared UMAP visualization

We will make a two-dimensional UMAP so that we can visualize our later cluster assignments. We run UMAP on the 20 PCA scores rather than on all SNP columns.

```{r run-umap, cache=TRUE}
set.seed(123)


umap_result <- umap(
  pca_scores,
  n_neighbors = 15,
  min_dist = 0.1,
  metric = "euclidean",
  verbose = FALSE
)

umap_coordinates <- umap_result$layout

df_plot <- df %>%
  mutate(
    UMAP1 = umap_coordinates[, 1],
    UMAP2 = umap_coordinates[, 2]
  ) %>%
  select(-starts_with("rs")) 
```

Before clustering, plot the UMAP and color the observations by their known geographic region.

```{r plot-region-umap}
ggplot(df_plot, aes(x = UMAP1, y = UMAP2, color = region)) +
  geom_point(alpha = 0.5, size = 1.5) +
  labs(
    title = "HGDP individuals",
    subtitle = "UMAP colored by known geographic region",
    color = "Region"
  ) +
  theme_minimal()
```

### Question 2

**Describe two patterns that you see in the UMAP. Which regions appear relatively distinct, and which overlap?**

*Answer: The UMAP method produces five distinct local neighborhoods and of these neighborhoods, two appear fairly homogenous, two appear mostly homogenous, and one is heterogenous. The regions that appear relatively distinct are Oceania, East Asia, Africa, and America. The ones that overlap appear to be Central South Asia, Middle East, and Europe.*


UMAP is useful for visualizing local neighborhoods, but it changes the geometry of the data. Remember that distances and cluster sizes in a two-dimensional UMAP should therefore be interpreted cautiously.

> **We will cluster the 20 PCA scores and then display the assignments on the UMAP. We will not use the two UMAP coordinates as the input to k-means.**

## Run k-means clustering to discover clusters in an unsupervised way

K-means begins with a number of clusters, $K$, chosen by the analyst. It then alternates between two steps:

1. Assign each observation to its nearest cluster center (centroid).
2. Recalculate each centroid as the mean position of the observations assigned to it.

The process repeats until the assignments stop changing.

Run k-means with $K=6$. `nstart = 25` tries 25 different random initial sets of centroids and keeps the solution with the smallest total within-cluster variation.

```{r run-kmeans}
set.seed(123)

km6 <- kmeans(
  pca_scores,
  centers = 6,
  nstart = 25
)

```



### Question 3

**Examine `km6$cluster`, `km6$centers`, and `km6$size`. What does each object contain? How many values or rows do you expect in each?**

```{r inspect-kmeans}
head(km6$cluster)
km6$centers
km6$size
```

*Answer: For the object km6, km6\$cluster is a vector of length 1107 containing a cluster label, an integer, for each observation. For km6\$centers, this contains the coordinates (20) for each centroid or cluster center, so we expect six rows with 20 values. The object km6\$size contains six values for how many observations belong to a cluster*

```{r}

# add the cluster to the data so we can plot it
# Find the component of km6 that will give you a column of cluster ids
df_plot <- df_plot %>%
  mutate(kmeans_cluster = factor(km6$cluster))
```


Now display the cluster assignments on the same UMAP.

```{r plot-kmeans-umap}
ggplot(df_plot, aes(x = UMAP1, y = UMAP2, color = kmeans_cluster)) +
  geom_point(alpha = 0.75, size = 1.5) +
  labs(
    title = "K-means clustering of HGDP individuals",
    subtitle = "K-means used 20 PCs; assignments are displayed on UMAP",
    color = "K-means cluster"
  ) +
  theme_minimal()
```

### Question 4

**Does each k-means cluster occupy a compact part of the UMAP? Identify one boundary or group of observations where the k-means assignments do not match what you would have drawn by eye.**

*Answer: In the visualization, four of the k-means clusters (2, 3, 4, and 6) occupy a compact region in UMAP space. For cluster 1 and 5, these do not appear to be compact as some points are in proximity to other clusters (4, and 6). One group of observations where the cluster assignment does not match what can be drawn by eye is those close to cluster 6 that are labeled as cluster 1 and 5. If drawing by eye, these should be included in cluster 6.*

### Question 5

**Why might clusters identified using PCA-transformed data not perfectly match the groups you see visually on the UMAP? Consider how PCA and UMAP represent the structure of the data.**

*Answer: PCA finds eigenvectors, or axes, that capture linear variance in a data set. UMAP captures the neighborhood structure of a data set in higher dimensions when projecting it to fewer dimensions, and, unlike PCA, it is a non-linear method. Given these differences, clusters identified with PCA-transformed data may not match groups seen visually with UMAP.*

## Compare clusters with known regions

K-means did not receive the `region` labels. We can now compare its data-driven assignments with those known labels.

```{r cluster-region-table}
cluster_region_counts <- df_plot %>%
  count(kmeans_cluster, region) %>%
  group_by(kmeans_cluster) %>%
  mutate(proportion = n / sum(n)) %>%
  ungroup()

cluster_region_counts
```

The following stacked bar plot shows the regional composition of each cluster.

```{r cluster-region-bars}
ggplot(
  cluster_region_counts,
  aes(x = kmeans_cluster, y = proportion, fill = region)
) +
  geom_col() +
  scale_y_continuous(labels = scales::percent) +
  labs(
    x = "K-means cluster",
    y = "Individuals in cluster",
    fill = "Region"
  ) +
  theme_minimal()
```

### Question 6

**How well do the six clusters correspond to the known geographic regions? Give one example of a relatively homogeneous cluster and one example of a cluster that contains individuals from multiple regions.**

*Answer: Of the six clusters, four correspond relatively well to known geographic regions (Oceania, Africa, East Asia, and America), while clusters 1 and 5 appear to be mixtures of regions. Cluster 2, labeled Oceania, appears homogeneous, while cluster 5 is a mixture of the Middle East, Europe, Central South Asia, and Africa.*

### Question 7

**If a k-means cluster contains several geographic regions, does that mean the algorithm failed? Explain why or why not.**

*Answer: If a k-means cluster contains several geographic regions, it does not mean that the algorithm failed. For this data set, k-means groups individuals based on their genetic composition, with no weight given to geography. A given region may contain multiple genotypes due to migration. Additionally, the selection of k will influence which regions are lumped together. In this data set, since k=6 and regions=7, some groups will contain several geographic regions.*


## How should we choose K?

K-means requires us to select $K$ before fitting the model. One diagnostic is the **total within-cluster sum of squares**: the squared distances between observations and their assigned centroids. Lower values mean that observations are, on average, closer to their centroids.

Calculate this value for $K=1$ through $K=12$.

```{r elbow-calculation, cache=TRUE}
set.seed(123)

# Calculate for k=1 through k=12
elbow_results <- tibble(K = 1:12) %>%
  mutate(
    total_within_ss = map_dbl(
      K,
      ~ kmeans(pca_scores, centers = .x, nstart = 25)$tot.withinss
    )
  )

ggplot(elbow_results, aes(x = K, y = total_within_ss)) +
  geom_line() +
  geom_point() +
  scale_x_continuous(breaks = 1:12) +
  labs(
    title = "Elbow plot",
    x = "Number of clusters (K)",
    y = "Total within-cluster sum of squares"
  ) +
  theme_minimal()
```

Adding clusters will always reduce the total within-cluster sum of squares. We therefore look for an **elbow**: a point after which increasing $K$ gives progressively smaller improvements. The elbow is a guide, not an automatic or uniquely correct answer.

### Question 8

**Where do you see a possible elbow? Would you choose $K=6$ based on this plot alone? Explain your reasoning.**

*Answer: Visually, a possible elbow is at $k=4$. Based on the plot alone, I would not choose k=6, as this value is past the elbow, and going from k=4 to k=6 does not significantly improve the total within-cluster sum of squares, meaning the points are relatively stable around the centroids.*


## Explore a different value of K

Choose one other value of $K$, rerun k-means, and display the new assignments on the UMAP. Replace `???` with your selected value.

```{r choose-k}
chosen_k <- 7

set.seed(123)

km_chosen <- kmeans(
  pca_scores,
  centers = chosen_k,
  nstart = 25
)

df_plot <- df_plot %>%
  mutate(chosen_cluster = factor(km_chosen$cluster))

ggplot(df_plot, aes(x = UMAP1, y = UMAP2, color = chosen_cluster)) +
  geom_point(alpha = 0.75, size = 1.5) +
  labs(color = "Cluster") +
  theme_minimal()
```

### Question 9

**What changed when you changed $K$? Did the new solution split a previously coherent group, combine distinct groups, or reveal a useful level of structure? Use a specific region of the UMAP in your answer.**

*Answer: Changing to $k=7$ split a previously coherent group in the bottom-left of the plot. In both the original UMAP plot (labeled by region) and the k-means plot, this group appears fairly homogeneous. As $k$ changes, this same group now appears to consist of two subpopulations.*

## Final Reflection

### Question 10

**Brainstorm a situation where clustering data could be useful. What would the observations and features be, and what might the resulting clusters reveal? (3 or 4 sentences) **

*Answer: Clustering may be useful for analyzing the bacterial microbiome composition in samples from different parts of corn plants (Zea mays). Observations would include sample type (e.g., rhizosphere, root, shoot, and leaf). Features are the relative abundance of bacterial taxa in each sample. Clustering may reveal whether different tissue types have similar bacterial microbiomes.*



## Take-home ideas

- K-means assigns every observation to one of $K$ clusters.
- The method defines a cluster as observations that are close to a shared centroid.
- The analyst must choose $K$, and several values may be scientifically useful.
- Known labels can help interpret clusters, but they were not used to construct them.
- UMAP is a display here; the clustering was performed on the 20-dimensional PCA representation.

In the next lesson, we will use **hierarchical clustering**. Instead of choosing $K$ first, we will build a tree of relationships and then decide where to cut it.
