---
title: "Exercise - Multiple Hypotheses and FDR"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We'll be working with our favorite gene expression data set again.
First, load up the `tidyverse` and our expression data.
```{r message=FALSE}
library(tidyverse)
load("Annotated_Results_LvV.RData")
```

And filter our data as before and add the `upregulated` column.
```{r}
shrinkLvV <- shrinkLvV %>% filter(!is.na(Entrez))
shrinkLvV <- shrinkLvV %>% mutate(upregulated = (logFC > 1) & (pvalue < 0.05))
```

Today, we'll be testing against a larger set of gene sets, from the Kyoto Encyclopedia of Genes and Genomes. Download the file Mm.c2.cp.kegg.v7.1.entrez.rds on d2l from  Datasets -> FuGeneExpression, and save it in your working directory. The we load the gene set definitions.
```{r}
db <- readRDS('Mm.c2.cp.kegg.v7.1.entrez.rds')
```

Now we'll calculate p-values for all of our tests.
To do that, we'll first define a function that calculates the p-value for a single gene set being overrepresented in the upregulated genes.
```{r}
enrichment_pval <- function(gs) {
  ul <- shrinkLvV %>% filter((Entrez %in% db[[gs]]) & upregulated) %>% nrow()
  ur <- shrinkLvV %>% filter((Entrez %in% db[[gs]]) & !upregulated) %>% nrow()
  ll <- shrinkLvV %>% filter(!(Entrez %in% db[[gs]]) & upregulated) %>% nrow()
  lr <- shrinkLvV %>% filter(!(Entrez %in% db[[gs]]) & !upregulated) %>% nrow()
  table <- matrix(c(ul,ll,ur,lr), nrow=2)
  return(fisher.test(table,alternative="greater")$p.value)
}
```
Then we use a `map` function to apply this function to all the gene sets
```{r}
all_pvals <- map_dbl(names(db), enrichment_pval)
```
And finally we store the results in a new tibble 
```{r}
all_tests <- tibble(set=names(db), p.value=all_pvals)
```

### Problem 1

***Plot a histogram of the p-values from all the tests.*** Hint: use the `geom_histogram` function in ggplot, with the x-coordinate given by `p.value`.

```{r}
ggplot(data=all_tests, aes(x=p.value)) +
  geom_histogram()
```

***What does the peak of tests near p=0 indicate?***


### Problem 2

Sort your `all_tests` data frame by `p.value` using
```{r}
all_tests <- all_tests %>% arrange(p.value)
```

Then use mutate to add a column to all_tests called Bonferroni that represents Bonferroni-adjusted p-values. ***Show your code.***

```{r echo=FALSE}
all_tests <- all_tests %>% 
  mutate(bonferroni = nrow(all_tests) * p.value)
```

***How many gene sets are statistically significant at a Bonferroni-adjusted p-value threshold of 0.05?***

***Of those gene sets, how many are expected to be false positives?***


### Problem 3

***Add a column to all_tests called `fdr` that contains Benjamini-Hochberg adjusted p-values.***
```{r}

```

(Hint: The expression `1:m` will create a vector of integers from `1` to `m`. You may want to create an intermediate `rank` column to help you calculate `fdr`.)

***How many genes sets are statistically significant at a false discovery rate of 0.05?***

***Of those gene sets, how many are expected to be false positives?***

### Problem 4

***Print out the names of the gene sets with FDR < 0.05. Biologically, why might this set of genes be upregulated in mammary cells from lactating mice?*** (Hint: See the abstract of ref. 24 from the Fu et al. 2015 paper in the D2L Datasets folder to learn about the importance of lipids in lactation.)
