---
title: "Exercise - Multiple Hypotheses and FDR"
author: "Luis Alegría"
date: "`r Sys.Date()`"
output:
  pdf_document: default
  html_document: default
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We'll be working with our favorite gene expression data set again.
First, load up the `tidyverse` liand our expression data.
```{r message=FALSE}
library(tidyverse)
load("Annotated_Results_LvV.RData")
```

And filter our data as before and add the `upregulated` column.
```{r}
shrinkLvV <- shrinkLvV %>% filter(!is.na(Entrez))
shrinkLvV <- shrinkLvV %>% mutate(upregulated = (logFC > 1) & (pvalue < 0.05))
```

Today, we'll be testing against a larger set of gene sets, from the Kyoto Encyclopedia of Genes and Genomes. Download the file Mm.c2.cp.kegg.v7.1.entrez.rds on d2l from  Datasets -> FuGeneExpression, and save it in your working directory. The we load the gene set definitions.
```{r}
db <- readRDS('Mm.c2.cp.kegg.v7.1.entrez.rds')
```

Now we'll calculate p-values for all of our tests.
To do that, we'll first define a function that calculates the p-value for a single gene set being overrepresented in the upregulated genes.
```{r}
enrichment_pval <- function(gs) {
  ul <- shrinkLvV %>% filter((Entrez %in% db[[gs]]) & upregulated) %>% nrow()
  ur <- shrinkLvV %>% filter((Entrez %in% db[[gs]]) & !upregulated) %>% nrow()
  ll <- shrinkLvV %>% filter(!(Entrez %in% db[[gs]]) & upregulated) %>% nrow()
  lr <- shrinkLvV %>% filter(!(Entrez %in% db[[gs]]) & !upregulated) %>% nrow()
  table <- matrix(c(ul,ll,ur,lr), nrow=2)
  return(fisher.test(table,alternative="greater")$p.value)
}
```
Then we use a `map` function to apply this function to all the gene sets
```{r}
all_pvals <- map_dbl(names(db), enrichment_pval)
```
And finally we store the results in a new tibble 
```{r}
all_tests <- tibble(set=names(db), p.value=all_pvals)
```

### Problem 1

***Plot a histogram of the p-values from all the tests.*** Hint: use the `geom_histogram` function in ggplot, with the x-coordinate given by `p.value`.

```{r}
ggplot(data=all_tests, aes(x=p.value)) +
  geom_histogram()
```

***What does the peak of tests near p=0 indicate?***

*The peak of tests near p=0 indicates that there is gene sets for which the null hypothesis can be rejected, which is that there is no change in expression between a particular set of genes and the rest of the genes.*

### Problem 2

Sort your `all_tests` data frame by `p.value` using
```{r}
all_tests <- all_tests %>% arrange(p.value)
```

Then use mutate to add a column to all_tests called Bonferroni that represents Bonferroni-adjusted p-values. ***Show your code.***

```{r}
all_tests <- all_tests %>% 
  mutate(bonferroni = nrow(all_tests) * p.value)
```

***How many gene sets are statistically significant at a Bonferroni-adjusted p-value threshold of 0.05?***

```{r}
all_tests %>% 
  filter(bonferroni < 0.05) %>% 
  nrow()
```

*Seven gene sets are statistically significant using a Bonferroni-adjusted p-value threshold of 0.05.*

***Of those gene sets, how many are expected to be false positives?***

*Of those gene sets, less than 0.05 are expected to be false positives.*

### Problem 3

***Add a column to all_tests called `fdr` that contains Benjamini-Hochberg adjusted p-values.***
```{r}
all_tests <- all_tests %>%
  mutate(rank = 1:nrow(all_tests)) %>%
  mutate(fdr = p.value * nrow(all_tests) / rank)
```

(Hint: The expression `1:m` will create a vector of integers from `1` to `m`. You may want to create an intermediate `rank` column to help you calculate `fdr`.)

***How many genes sets are statistically significant at a false discovery rate of 0.05?***

```{r}
all_tests %>% 
  filter(fdr < 0.05) %>% 
  nrow()
```

*Ten gene sets are statistically significant at a false discovery rate of 0.05.*

***Of those gene sets, how many are expected to be false positives?***

*Of these gene sets, around 5% are expected to be false postives.*

### Problem 4

***Print out the names of the gene sets with FDR < 0.05. Biologically, why might this set of genes be upregulated in mammary cells from lactating mice?*** (Hint: See the abstract of ref. 24 from the Fu et al. 2015 paper in the D2L Datasets folder to learn about the importance of lipids in lactation.)

```{r}
all_tests %>% 
  filter(fdr < 0.05) %>% 
  select(set)
```

_The names of the gene sets with FDR < 0.05 indicate that they are involved in lipid metabolism. Lipids are synthesized in large quantities in the mammary glands of lactating mice (1). Thus, biologically, gene sets involved in lipid metabolism are upregulated in lactating mice. **Source: (1) Rudolph, Michael C et al. “Metabolic regulation in the lactating mammary gland: a lipid synthesizing machine.” Physiological genomics vol. 28,3 (2007): 323-36. doi:10.1152/physiolgenomics.00020.2006**_ 